---
title: "Final Project Exploratory Data Analysis"
author: "Em Tallman, Jirat Rymparsurat, Martin Truong, Tyler Takeuchi"
output: html_document
date: "5/22/22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("knitr")
source("summary.r")
source("scatter and lines for final project2.0.R")
source("State Rate Chart.R")
source("map.r")
```

## Introduction

Our exploratory analysis will try to answer "How much has suicide changed through the years?", "What is the amount of suicide percent to actual deaths?", and "What states have the highest rates of suicide?"

We are focusing mostly on suicides in relation to other data.

## Summary Information

```{r summary, include=FALSE}
#print(summary)
# select which values we want to show since showing the entire table takes up
# a lot of space
```

Looking through our data, we found that the most common age group to commit suicide was `r summary$most_common_age` and the most common sex was `r summary$most_common_sex`. Looking at all suicide minimums per state (grouped by year), we found that the absolute lowest minimum came from `r summary$lowest_min_deaths_state` (Vermont), in the year `r summary$lowest_min_deaths_year`. Likewise, looking at maximums per state, we found that the absolute highest maximum came from `r summary$highest_max_deaths_state` (Rhode Island) in the year `r summary$highest_max_deaths_year`. In the case of human resource workers, in the year 2016 we found that there were (per 100k individuals): `r summary$psychiatrists_per_100k` psychiatrists, `r summary$psychologists_per_100k` psychologists, `r summary$soc_workers_per_100k` social workers, and `r summary$nurses_per_100k` nurses. 

## Table of Aggregated Data

```{r}
# code to show table here

```

### Chart 1 -

```{r}
# code to show chart

```

### Chart 2 - Number of Suicides Per year (2000-2015)

The purpose of including this scatter plot with lines is to show the change over years. The data shows an increase as the years go on. This could be due to the increase in data collection capabilities, or just a changing social world with more internet access as well. Also with the increase in internet access it also means increase in information.

```{r echo=FALSE}
<<<<<<< HEAD

suicidebyyearUSA <- filter(
  WorldSuicide,
  country == "United States of America") %>%
  filter(year >= 2000)%>% 
  group_by(year) %>% 
  summarize(suicides_no = sum(suicides_no))%>%
  na.omit(WorldSuicide)

scatterline <- ggplot(suicidebyyearUSA, aes(x = year, y = suicides_no)) +
  geom_point() +
  labs(x="Years", y="No.of Suicides", title = "No. of Suicides By Year") +
  geom_line(color="gray28") +
  geom_point(shape=23, color="#000000", fill="#261bcc", size=4)

  
print(scatterline)
=======
scatterline
>>>>>>> 6104fe5b0c8ad4809c38d195f8c95d56f7e36f04

```

### Chart 3 - US Map of Suicide Death Rates and Total Deaths

The purpose of including the 2 maps is to compare the average death rate by state per year to the average total deaths per year. This shows how a higher average death rate from suicide does not correlate to a higher average total death rate. The reason for this is that states with a lower population will have a higher average suicide death rate compared to states with a higher population because if they have the same amount of total deaths from suicide, then the state with the lower population will have a higher death rate. For example, California has the highest average total death, but has one of the lowest average death rates by suicide since it has one of the highest state population in the country.

```{r echo=FALSE}
avg_rate_map
avg_death_map

```
