---
title: "Final Project Exploratory Data Analysis"
author: "Em Tallman, Jirat Rymparsurat, Martin Truong, Tyler Takeuchi"
output: html_document
date: "5/22/22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("knitr")
source("summary.r")
source("scatter and lines for final project2.0.R")
source("State Rate Chart.R")
source("map.r")
```

## Introduction

Our exploratory analysis will try to answer "How much has suicide changed through the years?", "What is the amount of suicide percent to actual deaths?", and "What states have the highest rates of suicide?"

We are focusing mostly on suicides in relation to other data.

## Summary Information

```{r summary}
print(summary)
# select which values we want to show since showing the entire table takes up
# a lot of space
```

Some values that we calculated are ... These values show ...

## Table of Aggregated Data

```{r}
# code to show table here

```

### Chart 1 -

```{r}
# code to show chart

```

### Chart 2 - Number of Suicides Per year (2000-2015)

The purpose of including this scatter plot with lines is to show the change over years. The data shows an increase as the years go on. This could be due to the increase in data collection capabilities, or just a changing social world with more internet access as well. Also with the increase in internet access it also means increase in information.

```{r echo=FALSE}
<<<<<<< HEAD

suicidebyyearUSA <- filter(
  WorldSuicide,
  country == "United States of America") %>%
  filter(year >= 2000)%>% 
  group_by(year) %>% 
  summarize(suicides_no = sum(suicides_no))%>%
  na.omit(WorldSuicide)

scatterline <- ggplot(suicidebyyearUSA, aes(x = year, y = suicides_no)) +
  geom_point() +
  labs(x="Years", y="No.of Suicides", title = "No. of Suicides By Year") +
  geom_line(color="gray28") +
  geom_point(shape=23, color="#000000", fill="#261bcc", size=4)

  
print(scatterline)
=======
scatterline
>>>>>>> 6104fe5b0c8ad4809c38d195f8c95d56f7e36f04

```

### Chart 3 - US Map of Suicide Death Rates and Total Deaths

The purpose of including the 2 maps is to compare the average death rate by state per year to the average total deaths per year. This shows how a higher average death rate from suicide does not correlate to a higher average total death rate. The reason for this is that states with a lower population will have a higher average suicide death rate compared to states with a higher population because if they have the same amount of total deaths from suicide, then the state with the lower population will have a higher death rate. For example, California has the highest average total death, but has one of the lowest average death rates by suicide since it has one of the highest state population in the country.

```{r echo=FALSE}
avg_rate_map
avg_death_map

```
